Fast gravitational wave parameter estimation with no-compromise

### 2023-01-22 src/tex/ms.tex
Address max's comment up to section 3

\te{Do we also already here want to make the point that it really helps with memory issues of GPUs if the frequency array isn't too large?}

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems that
most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit locally-varying
correlations, since they assume a uniform global structure by construction
\cite{2017arXiv170102434B}.\footnote{Sampling algorithms that use the
information of higher order derivatives such as manifold-MALA and
Riemannian-HMC \cite{RMHMC} can in principle handle local correlations in the
target distribution; however, they often encounter instabilities when used in
real-life applications, so their use is rare practice.} Another example is
multi-modality: if there are multiple modes in the target distribution,
individual chains will likely be trapped in one mode and take an extremely long
time to transverse between the modes \cite{2018arXiv180803230M}.  This means
that the relative weights between modes will take much longer to estimate than
the shape of each mode.

@@

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
are a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|^{-1},
\end{align}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. For a detailed discussion of the algorithm, we refer the readers to
\cite{2019arXiv190809257K, 2019arXiv191202762P}.

% The flowMC algorithm
\mi{can we clarify what it means to learn the ``global landscape'', and how
this jives with the fact that norm flows are supposed to help with the
\emph{local} correlations?}
Working in tandem, gradient-based MCMC and normalizing flows can efficiently
explore posteriors with local and global correlations, as well as multiple
separate modes.  The scheme relies on iteratively using draws from the
gradient-based MCMC to train a normalizing flow, which is then itself used as a
proposal for another stage of MCMC sampling.

Concretely, we begin by producing initial training data for the normalizing
flow by running multiple independent chains of the gradient-based algorithm for
a fixed number of steps.  From the resulting pool of samples, the normalizing
flow can begin to learn the global landscape of the target distribution.
%
However, since the independent chains contain the same number of samples, the
relative weight assigned to each chains will not represent the true target
distribution (e.g., the relative importance of separate modes will not be
correctly calibrated). This is mitigated by a second stage of gradient-based
MCMC sampling that uses the distribution learned by the normalizing flow as a
\textit{proposal}.

@@

Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. We use the package
\textsc{flowMC} \cite{2022arXiv221106397W}. The pseudocode of the algorithm is
given in Algorithm \ref{alg:cap}.

@@

Modern \mi{hardware?} accelerators such as graphics processing units (GPUs) and
tensor processing units (TPUs) are designed to execute large-scale dense
computation. They are often much more cost-efficient than using many central
processing units (CPUs) when it comes to solving problems that can be benefited
from parallelization.  The downside of these accelerators compared to CPUs is
that they can only perform a more restricted set of operations and are often
less performant when dealing with serial problems. Parameter estimation with
MCMC is a serial problem since each new sample generated from a chain depends
on the last sample in the chain. This means that naively putting the problem on
an accelerator is more likely to harm performance than improve it.

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform models used in PE are approximants, they can be evaluated at any
given time or frequency independently \mi{is this true for SEOB, and for the
surrogate?}. This means computing a waveform can be trivially parallelized over
frequency bins. Together with the heterodyned likelihood, we can evaluate the
likelihood at $\mathcal{O}(10^7)$ different locations on an Nvidia A100 GPU
\mi{faster?}.  The high throughput of likelihood evaluations unlocks the
potential of the \texttt{flowMC} sampling algorithms.

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 seconds of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz.  We draw noise from the design
PSDs for the LIGO Hanford, LIGO Livingston and Virgo detectors \cite{} \mi{what
curves did you use?}.
For both injection and recovery, we make use of the \textsc{IMRPhenomD}
waveform \cite{Khan:2015jqa} via the fully-differentiable implementation
presented in the \textsc{ripple} package \cite{}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
\mi{cite}} for GW150914 and GW170817 respectively. Running on an Nvidia A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in
%\kw{\variable{data/} and \variable{script}}
.

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    GW150914 posterior computed by our code (\mi{blue}) and \textsc{Bilby} (\mi{orange}).
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    GW170817 posterior computed by our code (\mi{blue}) and \textsc{Bilby} (\mi{orange}).
}
\label{fig:GW170817}
\end{figure}

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
\mi{we should say this in the het like sect}
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.

>>>>

% \te{Do we also already here want to make the point that it really helps with memory issues of GPUs if the frequency array isn't too large?}

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems that
most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit
locally-varying correlations, since they assume a single mass matrix that does
not depend on the location of the chain by construction
\cite{2017arXiv170102434B}. \footnote{Sampling algorithms that use the
information of higher order derivatives such as manifold-MALA and Riemannian-HMC
\cite{RMHMC} can in principle handle local correlations in the target
distribution; however, they often encounter instabilities when used in real-life
applications, so their use is a rare practice.} Another example is multi-modality:
if there are multiple modes in the target distribution, individual chains will
likely be trapped in one mode and take an extremely long time to transverse
between the modes \cite{2018arXiv180803230M}.  This means that the relative
weights between modes will take much longer to estimate than the shape of each
mode.

@@

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
is a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|^{-1},
\end{align}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. For a detailed discussion of the algorithm, we refer the readers to
\cite{2019arXiv190809257K, 2019arXiv191202762P}.

% The flowMC algorithm
Working in tandem, gradient-based MCMC and normalizing flows can efficiently
explore posteriors with local and global correlations, as well as multiple
separate modes.  The scheme relies on iteratively using draws from the
gradient-based MCMC to train a normalizing flow, which is then itself used as a
proposal for another stage of MCMC sampling.

Concretely, we begin by producing initial training data for the normalizing
flow by running multiple independent chains of the gradient-based algorithm for
a fixed number of steps.  From the resulting pool of samples, the normalizing
flow can begin to learn the landscape of the target distribution.
However, since the independent chains contain the same number of samples, the
relative weight assigned to each chain will not represent the true target
distribution (e.g., the relative importance of separate modes will not be
correctly calibrated). This is mitigated by a second stage of gradient-based
MCMC sampling that uses the distribution learned by the normalizing flow as a
\textit{proposal}.

@@

Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. We use the package
\texttt{flowMC} \cite{2022arXiv221106397W}. The pseudocode of the algorithm is
given in Algorithm \ref{alg:cap}.

@@

Modern hardwares such as graphics processing units (GPUs) and
tensor processing units (TPUs) are designed to execute large-scale dense
computation. They are often much more cost-efficient than using many central
processing units (CPUs) when it comes to solving problems that can be benefited
from parallelization.  The downside of these accelerators compared to CPUs is
that they can only perform a more restricted set of operations and are often
less performant when dealing with serial problems. Parameter estimation with
MCMC is a serial problem since each new sample generated from a chain depends
on the last sample in the chain. This means that naively putting the problem on
an accelerator is more likely to harm performance than improve it.

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform model we use can be evaluated at any given time or frequency
independently, this means computing a waveform can be trivially parallelized
over frequency bins. Together with the heterodyned likelihood, we can evaluate
the likelihood at $\mathcal{O}(10^7)$ different locations on an Nvidia A100 GPU.
The high throughput of likelihood evaluations unlocks the potential of the
\texttt{flowMC} sampling algorithms.

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 seconds of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz.  We draw noise from the design
PSDs for the LIGO Hanford, LIGO Livingston
(\textsc{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\textsc{SimNoisePSDAdvVirgo}) detectors \cite{}.
For both injection and recovery, we make use of the \textsc{IMRPhenomD} waveform
\cite{Khan:2015jqa} via the fully-differentiable implementation presented in the
\textsc{ripple} package \cite{}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
\cite{arviz_2019}} for GW150914 and GW170817 respectively. Running on an Nvidia A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in
%\kw{\variable{data/} and \variable{script}}
.

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW170817}
\end{figure}

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.
